---
title: "Markdown for run_analysis.R"
output: html_document
---




In the code below, lines 11-14, we first create a folder in the working directory called **projectdata**, and then download and unzip the files associated with the project using the provided URL.
```{r}
if(!file.exists("./projectdata")){dir.create("./projectdata")}
fileURL<- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
download.file(fileURL,destfile="./projectdata/Data.zip",method="curl")
unzip(zipfile="./projectdata/Data.zip",exdir="./projectdata")
```

Now, we begin reading the various tables in.  Below I am reading the tables on the files of the 'test' dataset from the "test" folder.  I am giving them handles like 'ytest', 'xtest', 'subtest' to keep things simple, so I can backtrack easily and refer to my raw data in the "projectfolder", if I need to down the road:
```{r}
ytest<- read.table("projectdata/UCI HAR Dataset/test/y_test.txt", header=FALSE)
xtest<- read.table("projectdata/UCI HAR Dataset/test/X_test.txt", header=FALSE)
subtest<- read.table("projectdata/UCI HAR Dataset/test/subject_test.txt", header = FALSE)

```

Here, I do the same with the files in the "train" folder:
```{r}
ytrain<- read.table("projectdata/UCI HAR Dataset/train/y_train.txt", header=FALSE)
xtrain<- read.table("projectdata/UCI HAR Dataset/train/X_train.txt", header=FALSE)
subtrain<- read.table("projectdata/UCI HAR Dataset/train/subject_train.txt", header = FALSE)
```

And finally, I also read in the files that I am going to use as variable names for the columns of my data frame, and as activity labels for the rows in my activity column:
```{r}
columnnames<- read.table("projectdata/UCI HAR Dataset/features.txt", header=FALSE)
activitynames<- read.table("projectdata/UCI HAR Dataset/activity_labels.txt", header=FALSE)
```

Now, I am ready as I have read in all the files and tables I need into my R console.  I can finally begin with the steps of my assignment.


**1.  Merges the training and the test sets to create one data set.**
For this, I merge the train and test files, by rows.  So I append **'xtest'** (line20) below **'xtrain'** (line 28).  
```{r}
mergeX<- rbind(xtrain, xtest)
```

And then I give names to the columns, using the values in column 2 from the table called "columnnames"  (line 34). *This step also partially fulfills step 4 of our assignment, which requires descriptive variable names.*
```{r}
colnames(mergeX)<- columnnames[,2] 
```

Do the same thing with the **'ytrain' 'ytest'** as well as  **'subtrain' 'subtest'** files.  Both the 'y' and 'sub' data frames are one column data frames. So I just give the column names using a character vector of length 1.
```{r}
mergey<- rbind(ytrain, ytest)
colnames(mergey)<- c("activity")
mergesub<- rbind(subtrain, subtest)
colnames(mergesub)<- c("subject")
```

If you check the three dataframes, **mergeX mergey mergesub** using the `str()` command, you will see that they all have 10299 rows.  We merge the three data frames into one, and we append them by columns:
```{r}
mergeall<- cbind(mergeX, mergesub, mergey)
```


**2. Extracts only the measurements on the mean and standard deviation for each measurement.**
For this, I use a very basic command that we use on data frames to select specific rows and columns --  `data.frame[rows, columns]`.  However, for the columns argument, I use the `grep()` function, and a variety of meta characters to make sure I extract all column names that have match the pattern **mean()** and **std()**.  I also make sure to extract the **activity** and **subject** columns because these provide information on the activities to which these measurements correspond, and the volunteers that participated in the study from whom these measurements were taken.  
Since I am essentially subsetting my merged dataset, I am calling it **subsetmerge**:
```{r}
subsetmerge<- mergeall[, grep("mean\\(\\)|std\\(\\)|activity|subject", colnames(mergeall))]
```


**3. Uses descriptive activity names to name the activities in the data set** 
The 'activities' actually correspond to the values in the 'y' files.  I have already added this column and named it "activity" above (line 55).  But I need to make them "descriptive, right now they are number values between 1 and 6.  So I am going to use the second column of the data frame **activitynames** (line 35).  I simply factor my "activity" column, the last column in data frame **subsetmerge** which currently has the class = integer.
```{r}
subsetmerge$activity <- factor(subsetmerge$activity, 1:6, as.character(activitynames$V2))
```
My "activity" column is now of class = factor, and 1:6 is now replaced with, in order, *walking, walking upstairs, walking downstairs, sitting, standing, laying.*


**4. Appropriately labels the data set with descriptive variable names.**
This was partially taken care of from the code above (line 49). But I need to clean up the column titles of typos, any unnecessary character, and make sure that the abbreviations don't become a guessing game for someone analyzing the data.  With `gsub()` command applied to the column names of **subsetmerge** data frame, I can substitute the "descriptive" variable names for the short forms, and remove unwanted characters like "()" that are currently at the end of all my column titles.
```{r}
colnames(subsetmerge)<-gsub("^tBody", "timeBody", colnames(subsetmerge))
colnames(subsetmerge)<-gsub("^tGravity", "timeGravity", colnames(subsetmerge))
colnames(subsetmerge)<-gsub("^fBody|^fBodyBody", "frequencyBody", colnames(subsetmerge))
colnames(subsetmerge)<-gsub("Acc", "Acceleration", colnames(subsetmerge))
colnames(subsetmerge)<-gsub("Gyro", "Gyroscope", colnames(subsetmerge))
colnames(subsetmerge)<-gsub("Mag", "Magnitude", colnames(subsetmerge))
colnames(subsetmerge)<-gsub("\\(\\)", "", colnames(subsetmerge))
```


**5. From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.**
Finally, I am at the last step of my assignment.  I have to make a second data frame using the variables (columns) and values (rows) in **subsetmerge**.  I can do that fairly easily if I install the *""dplyr package""* and load it using the `library()` command.

I simply group **subsetmerge** by "subject" first and "activity" second, storing the grouped data in **Groupmerge** -- `Groupmerge<- group_by(subsetmerge, subject, activity)`
I then use `summarize_each()` command on **Groupmerge**  with the *funs(mean)* as the second argument. 

My resulting Data Frame should be 180 rows and 68 columns, that gives just one value that averages  measurement for each of the 6 activities for each of the 30 volunteers.

DONE!  

